---
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---
<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/teaching/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

#Exercise Problems 3 

###Case Study: Survey of Students

This is the same data set we considered in Exercise Problems 1. The data is in **studentsurvey**

**Problem 1** What can you say about the relationship between Score and Gender? 

**Problem 2** What can you say about the relationship between Score and GPA? Find a 95% interval estimate for a student with a GPA of 2.5. Is this an interpolation or an extrapolation? 
  
**Problem 3** What can you say about the relationship between Score and Distance? Find a 99% interval estimate for a student who lives 1.5 miles from the school. Is this an interpolation or an extrapolation?
    
**Problem 4** What can you say about the relationship between Score and Age? Find a 90% interval estimate for the mean score of  21 year old students. Is this an interpolation or an extrapolation?

```{r}
attach(studentsurvey) 
```

**Problem 1** What can you say about the relationship between Score and Gender?

In problem 1 of the Exercise Problems 2 we ran the ANOVA and found a statistically significant difference between the scores of males and females. Because there are just two groups there is no reason to run tukey, but rerunning the same command gives us a 95% confidence interval for the difference in scores:

```{r}
oneway(Score, Gender) 
```

`r hl()$hr()`

**Problem2** What can you say about the relationship between Score and GPA? 

In problem 3 of the Exercise Problems 2 we  found a statistically significant correlation between Score and GPA. Let's find a good model. 

```{r}
slr(Score, GPA)
```

the residual vs fits plt and the normal plot looks good, so no problem with the assumptions. We find the model 

$$
\text{Score} = 3.12 + 1.33 \text{ GPA}
$$

Find a 95% interval estimate for a student with a GPA of 2.5. Is this an interpolation or an extrapolation?
```{r}
slr.predict(Score, GPA, newx=2.5, interval="PI")
```

so a 95% prediction interval  for a student with a GPA of 2.5 is (2.65, 10.24)

This is an interpolation because 2.5 is in the range of GPAs in the data set 
`r hl()$hr()`
**Problem 3** What can you say about the relationship between Score and Distance? 

In problem 5 of the Exercise Problems 2 we used a log transform on Distance. Doing so again yields 
```{r}
slr(Score, log(Distance + 1))
```
the residual vs fits plt and the normal plot looks good, so no problem with the assumptions. We find the model 

Score = 6.2 + 0.042 log(Distance+1) 

Find a 99% interval estimate for a student who lives 1.5 miles from the school. Is this an interpolation or an extrapolation?
```{r}
slr.predict(Score, log(Distance+1), 
        newx=log(1.5+1), interval="PI", conf.level = 99)
```
so a 95% prediction interval for a student who lives 1.5 miles from the school is (0.94, 11.54)

**Note** newx=log(1.5+1) because we have the predictor log(Distance+1).

This is an interpolation because 1.5 is in the range of Distances in the data set.
`r hl()$hr()`

**Problem 4 ** What can you say about the relationship between Score and Age? 

Before we saw that observation #220 is an outlier and removed it. We do the same now. Then
```{r}
slr(Age[-220], Score[-220])
```

Find a 90% interval estimate for the mean score of 21 year old students. Is this an interpolation or an extrapolation?

```{r}
slr.predict(Score[-220], Age[-220], 
            newx=21, interval="CI", conf.level = 90)
```
so a 90% confidence interval for the mean score of 21 year old students is (5.82, 6.44)

This is an interpolation because 21 is in the range of Ages in the data set. 


```{r, echo=FALSE}
detach(studentsurvey)
```
