---
output:
  pdf_document:
    fig_caption: no
  html_document: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/Resma3/Resma3.RData"))
require(knitr)
require(ggplot2)
require(grid)
```
`r hl()$basefontsize()`

<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

#Exercise Problems 2 

###Case Study: Survey of Students
This is a fake data set I made up for this exercise. It is supposed to be from a survey of students at some College. The data is in **studentsurvey**  

The variables are  
```{r}
colnames(studentsurvey)
```
Score is a combination of several questions designed to measure how "happy" they are to study at the College. A high number means more happiness. Distance is how far they live from the College.
Make sure your answers are **complete**.

**Problem 1** Is there a relationship between Score and Gender?

**Problem 2** Is there a relationship between Score and Year? 

**Problem 3** Is there a relationship between Score and GPA? 
  
**Problem 4** Is there a relationship between Score and Distance? 
    
**Problem 5 ** Is there a relationship between Score and Age?

**Problem 6** Is there a relationship between Gender and Major? 
    
**Problem 7 ** Is there a relationship between Gender and Age?

`r hl()$hr()`

`r hl()$hr()` 

```{r}
 attach(studentsurvey) 
```

**Problem 1** Is there a relationship between Score and Gender? 
Score is a quantitative variable and Gender is a categorical variable with two values, so this is a problem for ANOVA
```{r}
bplot(Score,Gender) 
```
The boxplot shows a slight difference between the genders. There are a few slight outliers, but they are no problem.
```{r}
stat.table(Score, Gender)
```
The test:
```{r}
oneway(Score, Gender)
```
1) Parameters of interest: means of scores of men and women  
2) Method of analysis: two sample t test  
3) Assumptions of Method: residuals have a normal distribution, or sample sizes are large enough 
4) $\alpha$ = 0.05  
5) Null hypothesis H~0~: $\mu_1~ = \mu_2$ (groups have the same mean)  
6) Alternative hypothesis H~a~: $\mu_1 \ne \mu_2$ (groups have different means)  
7) p value = 0.0033  
8) 0.0033 < 0.05, there is some evidence that the group means are not the same, the women tend to score higher than the men.

Assumptions:  

1) Normal residuals: normal plot looks ok.  
2) equal variance: 3*1.71 = 5.13 > 2.21, ok

`r hl()$hr()`

**Problem 2** Is there a relationship between Score and Year?
Score is a quantitative variable and Year is a categorical variable, so this is a problem for ANOVA
```{r}
bplot(Score, Year)
```
The table of summary statistics is
```{r}
stat.table(Score, Year)
```
The test:
```{r}
oneway(Score, Year)
```
1) Parameters of interest: means of scores by year  
2) Method of analysis: ANOVA   
3) Assumptions of Method: residuals have a normal distribution, or sample sizes are large enough  
4) $\alpha$ = 0.05   
5) Null hypothesis H~0~: $\mu_1~ = \mu_2 = \mu_3 = \mu_4$ (groups have the same mean)  
6) Alternative hypothesis H~a~: $\mu_i \ne \mu_j$ (some groups have different means)  
7) p value = 0.4855  
8) 0.4855 > 0.05, there is no evidence that the group means are not the same.

Assumptions:  

1) Normal residuals.  looks ok.  
2) equal variance: 3*1.8    = 5.4 >  2.2  

`r hl()$hr()`

**Problem 3** Is there a relationship between Score and GPA?
Score and GPA are both quantitative variables, so this is a problem for the **Pearson's Correlation Coefficient **.

The marginal plot shows some increase in Score as the GPA increases.
```{r, warning=FALSE}
mplot(Score, GPA)
```
There are a few slight outlier, no problem, though.

The test:
```{r, warning=FALSE}
pearson.cor(Score, GPA, rho.null = 0)
```
1) Parameter of interest: Pearson's correlation coefficient $\rho$  
2) Method of analysis: test based on normal theory  
3) Assumptions of Method: relationship is linear, there are no outliers  
4) $\alpha$ = 0.05  
5) H~0~: $\rho = 0$ (no relationship between Score and GPA)  
6) H~a~: $\rho \ne 0$ (some relationship between Score and GPA)  
7) p = 0.000  
8) 0 < 0.05, so we reject H~0~, there is a relationship between the Score and the GPA, apparently students with a higher GPA are happier.

`r hl()$hr()`

**Problem 4** Is there a relationship between Score and Distance? 
Score and Distance are both quantitative variables, so this is a problem for the **Pearson's Correlation Coefficient**.

The marginal plot shows some slight outliers. The  log transform fixes it.

```{r, warning=FALSE}
mplot(Score, Distance)
```
```{r, warning=FALSE}
mplot(Score, log(Distance+1))
```
some students apparently live very close to the school, Distance=0. For this reason we use log(Distance+1).

The test:
```{r, warning=FALSE}
pearson.cor(Score, log(Distance+1), rho.null = 0)
```
1) Parameter of interest: Pearson's correlation coefficient $\rho$  
2) Method of analysis: test based on normal theory  
3) Assumptions of Method: relationship is linear, there are no outliers  
4) $\alpha=0.05$  
5) H~0~: $\rho =0$ (no relationship between Score and Distance)   
6) H~a~: $\rho \ne 0$ (some relationship between Score and Distance) 
7) p = 0.7305  
8) $p>\alpha$, so we fail to reject H~0~, there is no evidence of a relationship between the Score and the Distance.

`r hl()$hr()`

**Problem 5 ** Is there a relationship between Score and Age?
The marginal plot shows one severe outlier:


```{r, warning=FALSE}
length(Score)
length(Age)
mplot(Score, Age)
```
Unfortunately the log transform does not help, so the only way to preceed is to eliminate the outlier.
```{r}
which(Age==max(Age))
```
The test:
```{r, warning=FALSE}
pearson.cor(Age[-220], Score[-220], rho.null = 0)
```
1) Parameter of interest: Pearson's correlation coefficient $\rho$  
2) Method of analysis: test based on normal theory  
3) Assumptions of Method: relationship is linear, there are no outliers  
4) $\alpha = 0.05$  
5) H~0~: $\rho=0$ (no relationship between Score and Distance)  
6) H~a~: $\rho \ne 0$ (some relationship between Score and Distance)  
7) p = 0.3052  
8) $p > \alpha$, so we fail to reject H~0~, there is no evidence of a relationship between the Score and the Age.

`r hl()$hr()`

**Problem 6 ** Is there a relationship between Gender and Major? 
Gender and Major are both categorical variables, so this is a problem for the **Chisquare test of Independence**.

```{r}
 table(Gender, Major) 
```
```{r}
chi.ind.test(table(Gender, Major))
```
1) Parameters of interest: measure of association  
2) Method of analysis: chi-square test of independence  
3) Assumptions of Method: all expected counts greater than 5  
4) $\alpha = 0.05$
5) H~0~: Classifications are independent = Gender and Major are independent  
6) H~a~: Classifications are dependent = Gender and Major are not independent  
7) p = 0.9664  
8) 0.9664 > 0.05, there is no evidence of a relationship between gender and major. 

`r hl()$hr()` 

**Problem 7 ** Is there a relationship between Gender and Age?
Age is a quantitative variable and Gender is a categorical variable with two values, so this is a problem for the **ANOVA**.

The boxplot shows a few serious outliers. One could try transformations, but because the outliers are from especially small and large observations these won't work. 

```{r}
bplot(Age, Gender)
bplot(log(Age), Gender)
```

**Solution 1: non-parametric method **  
The table of summary statistics is  
```{r}
stat.table(Age, Gender, Mean=FALSE )
```
Now
```{r}
kruskalwallis(Age, Gender)
```
1) Parameters of interest: 2 medians  
2) Method of analysis: Kruskal-Wallis  
3) Assumptions of Method: **none**  
4) $\alpha=0.05$   
5) Null hypothesis H~0~: M~1~=M~2~ (group medians are the same)  
6) Alternative hypothesis H~a~: M~1~$\ne$M~2~ (group medians are not the same)  
7) p value = 0.2611  
8) 0.2611 > 0.05, so we fail to reject H~0~, it seems the group medians are the same.  

**Solution 2: remove outlier **
```{r, eval=FALSE}
which(Age==max(Age))
```
```{r}
stat.table(Age[-220], Gender[-220])
oneway(Age[-220], Gender[-220])
```
1) Parameters of interest: means of age by gender  
2) Method of analysis: ANOVA  
3) Assumptions of Method: residuals have a normal distribution, or sample sizes are large enough 
4) $\alpha$ = 0.05  
5) Null hypothesis H~0~: $\mu$~1~ = $\mu$~2~  (groups have the same mean)  
6) Alternative hypothesis H~a~: $\mu$~1~ $\ne$ $\mu$~2~ (some groups have different means)  
7) p value = 0.3817
8) 0.3817 > 0.05, there is no evidence that the group means are not the same. 
Assumptions:
1) Normal residuals ok
Smallest sd:  1    Largest sd : 1 , 3*1>1, ok


```{r, echo=FALSE}
detach(studentsurvey)
```
