---
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---
<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

#Exercises 4

###Case Study: Survey of Students
This is the same data set we considerd in the previous exerceises. 

**Problem 1** Find the best model to predict Score from GPA and Distance.

**Problem 2** Analyse the data with score as the response and Gender and Years as factors (predictors).
  
**Problem 3** Find a 90% interval estimate for the score of a male student with a GPA of 2.15. Is this an interpolation or an extrapolation?

**Problem 4** Find a 95% interval estimate for the score of a 20 year old male student who lives 2 miles from school. (code the variable Gender, ignore the issue of parallel lines)

`r hl()$hr()`

`r hl()$hr()` 

```{r}
attach(studentsurvey) 
```

**Problem 1** Find the best model to predict Score from GPA and Distance.

In  Exercise Problems 3 we found a linear model in GPA and a log model in Distance, so let's try this:

```{r}
mlr(Score, cbind(GPA, log(Distance + 1)))
```

the plots look good, so no problem with the assumptions.

Notice that the name of the variable Distance is missing. If we want to fix that we can do this:

```{r}
X <- cbind(GPA, log(Distance + 1))
colnames(X) <- c("GPA", "log(Distance + 1)") 
mlr(Score, X)
```

Can we simplify the model?

```{r}
mallows(Score, X)
```

the smallest C~p~ is for the model with GPA only, so this is best. 

```{r}
slr(GPA, Score) 
```

`r hl()$hr()`

**Problem 2** Analyse the data with score as the response and Gender and Years as factors (predictors).

this is a twoway ANOVA problem. In Exercise Problems 2 we already looked at the boxplots and the summary statistics. Next we need to consider any possible interaction: 

In problem 3 of the Exercise Problems 2 we  found a statistically significant correlation between Score and GPA. Let's find a good model. 

```{r}
iplot(Score, Gender, Year)
```

there seems to be interaction here. Can we test for it? We would need repeated measurements:
```{r}
table(Gender, Year)
```

which we have. So

```{r}
twoway(Score, Gender, Year) 
```

the plots look good, so no problems with the assumptions. 

The test for interaction has p=0.5825 , so there is no evidence of interaction. We can refit with out it:

```{r}
twoway(Score, Gender, Year, with.interaction=FALSE) 
```

The test for Year has p=0.42, so Year does not effect score and so we can drop the term as well. We are now back to a oneway of Score by GPA, which we already analysed in Exercise Problems 2 and 3. 

`r hl()$hr()`

**Problem 3**  Find a 90% interval estimate for the score of a male student with a GPA of 2.15. Is this an interpolation or an extrapolation?

We have a quantitative response (Score), a quantitative predictor (GPA) and a catgorical predictor( Gender), so this is a regression problem with a dummy variable. 
 
```{r}
dlr(Score, GPA, Gender)
```

do we need the product term?

```{r}
independent.lines <- dlr(Score, GPA, Gender, return.model=TRUE)
parallel.lines <- dlr(Score, GPA, Gender, additive=TRUE,
                 return.model=TRUE)
nested.models.test(parallel.lines, independent.lines)
```

gives p=0.000, so the product term is indeed necessary.

Next we want a 90% interval estimate for the score of a male student with a GPA of 2.15:

```{r}
dlr.predict(Score, GPA, Gender, newx=2.15, 
            newz="Male", interval="PI", conf.level=90)
```

Is this an interpolation or an extrapolation?

```{r}
range(GPA[Gender=="Male"])
```

shows that 2.15 is in the range of data values for males, so this is an interpolation. 

`r hl()$hr()`

**Problem 4** Find a 95% interval estimate for the score of a 20 year old male student who lives 2 miles from school. (code the variable Gender, ignore the issue of parallel lines)

As before we need to remove observation #220 from the data set and we need to use the log tranform on Distance:

```{r}
X <- data.frame(Distance=log(Distance[-220]+1), Age=Age[-220], Gender=ifelse(Gender[-220]=="Male", 1, 0))
mlr(Score[-220] , X) 
```

When doing prediction there is usually no reason to simplify the model, so we won't use Mallows C~p~. Now

```{r}
newx <- cbind(Distance=log(2)+1, Age=20, Gender=1)
mlr.predict(Score[-220], X, newx=newx, interval="PI")
```


```{r, echo=FALSE}
detach(studentsurvey)
```
