---
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---
<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

#Nonlinear Regression Models

###Case Study: Fabric Wear

Results from an experiment designed to determine how much the speed of a washing machine effects the wear on a new fabric. The machine was run at 5 different speeds (measured in rpm) and with six pieces of fabric each.

```{r}
head(fabricwear)
```

The scatterplot of wear by speed shows a strong but non-linear relationship:

```{r}
attach(fabricwear)
splot(Wear, Speed, add.line=1)
```

How strong is a difficult question, because Pearson's correlation coefficient won't work here. If we tried slr we would see in the residual vs fits plot that there is a problem with the assumption of a linear model:

```{r}
slr(Wear, Speed)
```

So the question is: how do fit models other than straight lines?

There are two basic things we can try. The first is something we have already done, namely the **log transformation**

```{r, warning=FALSE}
splot(Wear, log(Speed), add.line=1)
splot(log(Wear), log(Speed), add.line=1)
splot(log(Wear), Speed, add.line=1)
```

unfortunately non of these looks very good

Some of these have names:

- log(y) vs. x is called an **exponential model**

- log(y) vs. log(x) is called a **power model**

The other solution to our problem is to fit a **Polynomial Model**:

**Linear** $y=\beta_0+\beta_1 x$

**Quadratic** $y=\beta_0+\beta_1 x+\beta_2 x^2$

**Cubic** $y=\beta_0+\beta_1 x+\beta_2 x^2+\beta_3 x^3$

and so on

How do we fit such a model? We can simply use the same routine  with the extra argument *polydeg=...*. For example for the quadratic model we do

```{r}
slr(Wear, Speed, polydeg=2)
```

What does such a curve look like? To draw the fitted line plot, that is the scatterplot with the fitted curve, just use

```{r}
flplot(Wear, Speed, polydeg=2)
```

This routine also does the log transform models:

```{r}
flplot(Wear, Speed, logy=TRUE)  
```

Similarly use flplot(Wear, Speed, logx=TRUE) or flplot(Wear, Speed, logx=TRUE, logy=TRUE) for the other log transforms

**Note**
There are two big differences in the way transformations and polynomial models work: 

- if we do a transformation we **replace** an old variable with a new one, if we do a polynomial model we **add** a new predictor to the model.

- we might transform the response, but a polynomial model is always a polynomial in the predictor, never the response. 

**Mathematical Features of these Models**

What "shapes" can we fit with these models?

- Transformations might work if the relationship between x and y is **monotone**, that is in the scatterplot the dots either go up or down but never turn around.

- Polynomial models usually do turn around, quadratic models once, cubic models twice and so on. Sometimes this is not apparent because we only see the graph before the turn-around happens. 
<hr>

###Prediction

Again we can use the slr.predict command to do prediction, but there are some things we need to be careful with:

**Transformations**

if we use a log transformation on the predictor we have to use the log transformation also on the newx: 

```{r}
slr.predict(Wear, log(Speed), newx=log(150))
```

if we use a log transformation on the response we are getting an estimate of the log of the response. To get back to the original we can do this:

```{r}
slr(log(Wear), Speed) 
```

so we have the equation

$$
\log(\text{Wear})  = 2.335 + 0.008 \text{Speed} 
$$

and now we can get an estimate with 

```{r}
exp(2.335 + 0.008*150)
```

All of this works ONLY for point estimation, interval estimation is much harder and needs to be done by an expert!

In contrast, prediction using polynomials works perfectly fine as is:

```{r}
slr.predict(Wear, Speed, newx=150, polydeg=2, 
            interval="PI", conf.level= 90)
```

If you are not sure that you got the right answer, here is a quick sanity check: draw the scatterplot and do a visual guess of y.

**Example**: say we want to use the power model and predict the Wear for Speed=150:

```{r}
slr.predict(log(Wear), log(Speed), newx=150)
```

Notice that I should have written newx=log(150).

Now if I draw the fitted line plot:

```{r}
splot(Wear, Speed)
```

it is clear that if x=150 y should be be around 32 or so, not 165!

So I better try again:

```{r}
slr.predict(log(Wear), log(Speed), newx=log(150))
```

and that's about right because

```{r}
log(32)
```


```{r, echo=FALSE}
detach(fabricwear)
```
